Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-22 manuscript/climate_sensitivity_draft.Rmd
Updated with coauthor feedback to end of paper
#65

Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volume provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

@@

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] **see comment 47** made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time. At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.  Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiolgically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015]. Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).

**see comment 48**
Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height. As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits. If vertical trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**; **see comment 49**), follow the general pattern of increasing leaf drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). Although there were not significant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig. 1, Table 4). Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019]. **see comment 50** In 1977, which was the weakest drought in terms of PDSI (Fig. S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015. In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted. Thus, responses differed with the nature of each drought. Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites--and across more droughts at a single site--would be of great value to elucidating the mechanisms through which drought characteristics interact with driver variables to shape tree growth responses.

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age. Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether and how drought tolerance changes as forests age. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

### Acknowledgements
Thanks to all researchers who helped to collect the data used here, in particular Jennifer C. McGarvey, Jonathan R. Thomspon, and Victoria Meakem for original collection and processing of cores. Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements and to Edward Brzostek's lab for collaboration on leaf sampling. ** others??** This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper. accurate here?** **coauthors, other funding?**)

>>>>

Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volum provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

@@

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] **see comment 47** made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time. At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.  Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiolgically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015]. Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).

**see comment 48; the first paragraph was condensed here, but unsure where this should be put now**
Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). Although there were not significant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig. 1, Table 4). Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019]. **see comment 49** In 1977, which was the weakest drought in terms of PDSI (Fig. S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015. In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted (Fig. S2). Thus, responses differed with the nature of each drought. Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites--and across more droughts at a single site--would be of great value to elucidating the mechanisms through which drought characteristics interact with driver variables to shape tree growth responses.

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young, non-dense forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age. Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

### Acknowledgements **see comment 51**
Thanks to all researchers who helped to collect the data used here, in particular Jennifer C. McGarvey, Jonathan R. Thomspon, and Victoria Meakem for original collection and processing of cores. Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection. This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper. accurate here?** **coauthors, other funding?**)
